---
title: "Post-Pandemic Expansion: How Is It Different?"
description: 
  Comparing The Current Recovery to Past Expansions
author:
  - name: Philipp Maier
    url: https://twitter.com/EconomicsShiny
date: 2021-11-07
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=15, fig.height=10) 

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop("Package not found")
  }
}

pkgTest("fredr")
pkgTest("RColorBrewer")
pkgTest("reshape")
pkgTest("quantmod")
pkgTest("zoo")
pkgTest("dplyr")
pkgTest("forecast")
pkgTest("lubridate")
pkgTest("httr")
pkgTest("jsonlite")
pkgTest("rtweet")
pkgTest("httr")
pkgTest("tis")
pkgTest("tm")

## Credentials

FRED_API_KEY <- "4a91875abccae0ee89966c96d8760b60"
fredr_set_key(FRED_API_KEY)


# Data Download

Data.US <- c("US.GDP.Real"="GDPC1", 
             "US.GDP.PCE"="PCECC96",
             "US.GDP.PCE.Goods"="DGDSRX1Q020SBEA",
             "US.GDP.PCE.DurableGoods"="PCDGCC96",
             "US.GDP.PCE.NondurableGoods"="PCNDGC96",
             "US.GDP.PCE.Services"="PCESVC96",
             "US.GDP.Investment"="GPDIC1",
             "US.GDP.Investment.Fixed"="FPIC1",
             "US.GDP.Investment.Nonresidential"="PNFIC1",
             "US.GDP.Investment.Residential"="PRFIC1",
             "US.GDP.Inventories"="CBIC1",
             "US.GDP.NetExports"="NETEXC",
             "US.GDP.Government"="GCEC1",
             "US.GDP.Government.Federal"="FGCEC1",
             "US.GDP.Government.StateLocal"="SLCEC1",
             "US.GDP.FinalSales"="FINSLC1",
             "US.Survey.Empire"="GACDINA066MNFRBNY", 
             "US.IP"= "INDPRO", 
             "US.Payroll"="PAYEMS",
             "US.IP.CapacityUtilization" = "TCU",
             "US.Unemployment"="UNRATE", 
             "US.Unemployment.U1" = "U1RATE", 
             "US.Unemployment.U6" = "U6RATE", 
             "US.Unemployment.U2" = "U2RATE",
             "US.Unemployment.PartTimeEconomicReasons" = "LNS12032194", 
             "US.Unemployment.PartTimeNonEconomicReasons" = "LNS12032200",
             "US.Unemployment.MarginallyAttached" = "LNU05026642",
             "US.Unemployment.ParticipationRate"="CIVPART",
             "US.Unemployment.EmploymentToPopulation"="EMRATIO",
             "US.Activity.ChicagoFed.Employment" = "EUANDH",
             "US.Activity.ChicagoFed" = "CFNAI",
             "US.Activity.PhillyFed.Current" = "USPHCI",
             "US.Activity.NYFed.Current" = "GACDISA066MSFRBNY",
             "US.Activity.NYFed.Leading" = "GAFDISA066MSFRBNY",
             "US.Activity.RetailSales" = "RSAFS",
             "US.Activity.RetailSalesExAuto" = "RSFSXMV",
             "US.Activity.ADP" = "NPPTTL",
             "US.Activity.InitialClaims" = "ICSA",
             "US.Activity.ContinuedClaims.4W.MA" = "CC4WSA",
             "US.Activity.ContinuedClaims" = "CCSA",
             "US.JOLTS.QuitsRate" = "JTSQUR",
             "US.JOLTS.HireRate" = "JTSHIR",
             "US.JOLTS.JobOpeningsRate" ="JTSJOR",
             "US.Unemployment.WageGrowth" = "CES0500000003",
             
             "US.CPI.Headline"="CPIAUCSL", 
             "US.CPI.Core"="CPILFENS",
             
             "US.CPI.Headline.mm.yy"="CPIAUCSL", 
             "US.CPI.Core.mm.yy"="CPILFENS",
             "US.CPI.AlcoholicBeverages.mm.yy" = "CUSR0000SEFW",
             "US.CPI.AlcoholicBeveragesAway.mm.yy" = "CUSR0000SEFX",
             "US.CPI.FoodBeverages.mm.yy" = "CPIFABSL",
             "US.CPI.Housing.mm.yy" = "CPIHOSSL",
             "US.CPI.Apparel.mm.yy" = "CPIAPPSL",
             "US.CPI.Medical.mm.yy" = "CPIMEDSL",
             "US.CPI.Recreation.mm.yy" = "CPIRECSL",
             "US.CPI.Education.mm.yy" = "CPIEDUSL",
             "US.CPI.Other.mm.yy" = "CPIOGSSL",
             
             
             "US.SOV.1Y"="DGS1", 
             "US.SOV.2Y"="DGS2", 
             "US.SOV.3Y"="DGS3", 
             "US.SOV.5Y"="DGS5", 
             "US.SOV.7Y"="DGS7", 
             "US.SOV.10Y"="DGS10", 
             "US.SOV.20Y"="DGS20", 
             "US.SOV.30Y"="DGS30", 

             "US.Transportation.Rail" = "RAILFRTINTERMODALD11",
             "US.Transportation.Railpassenger" = "RAILPMD11",
             "US.Transportation.Air" = "LOADFACTORD11",
             "US.Transportation.Airtraffic.Passenger" = "ASMD11",
             "US.Transportation.Airtraffic.International" = "ASMI",
             "US.Transportation.Index" = "TSIFRGHT",
             "US.Transportation.PublicTransit" = "TRANSITD11",
             
             "US.HouseholdDebt" = "HDTGPDUSQ163N",
             "US.Housing.NewPrivateHousingStarts"="HOUSTNSA",
             "US.Housing.NewPrivateHousingPermits"="PERMIT",
             "US.Housing.NewPrivateHousingConstruction"="UNDCONTNSA",
             "US.Housing.NewPrivateHousingCompleted"="COMPUTSA",
             "US.Housing.NewPrivate1UnitCompleted"="COMPU1USA",
             "US.Housing.NewPrivate2UnitCompleted"="COMPU24USA",
             "US.Housing.NewPrivate5UnitCompleted"="COMPU5MUSA",
             "US.Housing.30YMortgageRate"="MORTGAGE30US",
             "US.Housing.NewHomeSales" = "HSN1F",
             "US.Housing.ExistingHomeSales" = "EXHOSLUSM495S",
             "US.Housing.MonthlySupply"="MSACSR",
             "US.Housing.AllTransactionsPriceIndex"="USSTHPI",
             "US.Housing.CaseShiller" = "CSUSHPINSA",
             
             "US.Auto.Autosales" = "ALTSALES",
             "US.Auto.LightAutos" = "LAUTOSA",
             "US.Auto.LightTrucks" = "LTRUCKSA",
             "US.Auto.AutosDomestic" = "DAUTOSAAR",
             "US.Auto.AutosForeign" = "FAUTOSAAR",
             
             "US.Auto.InventorySalesRatio" = "AISRSA",
             "US.Auto.MilesTraveled" = "M12MTVUSM227NFWA",
             
             "US.H8.BanksCredit.Securities" = "SBCACBW027SBOG",
             "US.H8.BanksCredit.LoansLeases" = "TOTLL",
             "US.H8.BanksCredit.Allowance" = "ALLACBW027SBOG",
             "US.H8.InterBankLoans" = "LCBACBW027SBOG",
             "US.H8.FedFundsSold" = "H8B3092NCBA",
             "US.H8.Cash" = "CASACBW027SBOG",
             "US.H8.TradingAssets" = "H8B3053NCBA",
             "US.H8.BanksCredit.Securities.Treasuries" = "TASACBW027SBOG",
             "US.H8.BanksCredit.Securities.Other" = "OSEACBW027SBOG",
             "US.H8.BanksCredit.CI" = "TOTCI",
             "US.H8.BanksCredit.RE" = "RELACBW027SBOG",
             "US.H8.BanksCredit.Consumer" = "CLSACBW027SBOG",
             "US.H8.BanksCredit.Other" = "AOLACBW027SBOG",
             
             "US.FDIC.NetChargeOffRateTotalLoans" = "QBPLNTLNNTCGOFFR",
             "US.FDIC.LoanLossProvisions" = "QBPQYLNLOSS",
             "US.FDIC.UnprofitableInstitutions" = "QBPQYNUMINSTUNPR",
             "US.FDIC.InstitutionsWithEarningsGain" = "QBPQYNUMINSTYY",
             
             "US.Banks.AutoLoansSecuritized.qq.yy" = "MVLOAS",
             "US.Banks.StudentLoans.qq.yy" = "SLOAS",
             
             "US.GasPrices" = "GASALLW",
             "US.EconomicPolicyUncertaintyIndex" = "USEPUINDXD",
             "US.Commodities.Oilprices" = "DCOILWTICO",
             
             "EU.GDP.Real"="CLVMNACSCAB1GQEA19",
             "UK.GDP.Real"="NAEXKP01GBQ652S",
             "CA.GDP.Real"="NAEXKP01CAQ189S",
             "JP.GDP.Real" = "NAEXKP01JPQ661S"
             
)


Data.Description <- data.frame( 
  Mnemonic = names(Data.US), 
  Code = Data.US,
  Description = "", 
  Frequeny = "",
  Units = "",
  LastUpdate = ""
  )
  
for ( idx in 1:nrow(Data.Description) ) {
  cat(paste("  - ", Data.Description[idx,1], "\n"))
  x = fredr_series(series_id = Data.Description[idx,2])
  Data.Description[idx, 3 ] = x$title
  Data.Description[idx, 4 ] = x$frequency
  Data.Description[idx, 5] = x$units
  Data.Description[idx, 6] = x$last_updated
}


## Misc. functions

misc.FREDdowload <- function(series) {
  
  if (!(exists(series))) {
    x = fredr( series_id = Data.Description[Data.Description$Mnemonic == series,2] )
    x = zoo(x$value, as.Date(x$date))
    if ((series == "US.H8.FedFundsSold") || (series == "US.H8.TradingAssets")) x = x / 1000
    if ( substr(series, (nchar(series)-5), nchar(series)) == ".mm.yy") x <- 100*log(x / stats::lag(x, -12))
    if ( substr(series, (nchar(series)-5), nchar(series)) == ".qq.yy") x <- 100*log(x / stats::lag(x, -4))
    if ( substr(series, (nchar(series)-2), nchar(series)) == ".qq") x <- 100*log(x / stats::lag(x, -1))
    
  } else assign("x", get(series))
  
  assign(series, x, envir = .GlobalEnv)
  return(x)
}


misc.NBER.Recessions <- function(){
  NBER.Recessions <- as.Date(as.character(t(nberDates())), format="%Y%m%d")
  Limits <- par('usr')
  for (idx in seq(1, length(NBER.Recessions), 2) ) {
    rect(NBER.Recessions[idx], Limits[3], NBER.Recessions[idx+1], Limits[4], col="#0000FF19", lty=0)
  }
}


Chart.Panel <- function(x, series) {
  
  chart.col   = brewer.pal(6, "Paired")[6]
  chart.ylim  = c(0.9*range(x, na.rm=TRUE)[1], 1.1*range(x, na.rm=TRUE)[2])
  chart.title = Data.Description[Data.Description$Mnemonic == series,3]
  
  plot(as.zoo(x), main = chart.title, 
       col = chart.col, 
       lwd = 3, 
       ylim = chart.ylim,
       ylab = Data.Description[grep(series, Data.Description[,1]),5],
       xlab = paste0("Period: ", year(index(x[1])), " - ", year(index(tail(x[1]))), " (shaded areas indicate U.S. recessions)"))
  
  points( index(tail(x, 1)), tail(x,1), col = chart.col, pch = 19, lwd = 5)
  abline(v = as.Date(paste0(seq( year(index(x[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
  misc.NBER.Recessions()
  
  text(index(tail(x,round(.15*length(x)))[1]), 1.1*tail(x,1), paste("Latest obs.\n", round(tail(x,1),2)), font = 2)
  grid()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  
}


Chart.Single <- function(series, periods, tweet = FALSE, tweet.text = ""){
  
  x = misc.FREDdowload(series = series)
  x = x[year(index(x))>(year(Sys.Date())-periods)]
  
  chart.col   = brewer.pal(6, "Paired")[6]
  chart.ylim  = c(0.9*range(x, na.rm=TRUE)[1], 1.1*range(x, na.rm=TRUE)[2])
  chart.title = Data.Description[Data.Description$Mnemonic == series,3]
  if (nchar(chart.title) > 25) chart.title = gsub(":", ":\n", chart.title)
  
  plot(as.zoo(x), main = chart.title, 
       #ylab = "", 
       col = chart.col, 
       lwd = 3, 
       ylim = chart.ylim,
       ylab = "Source: FRED, Federal Reserve Bank of St. Louis",
       xlab = paste0("Period: ", year(index(x[1])), " - ", year(index(tail(x,1))), " (shaded areas indicate U.S. recessions)"))
  points( index(tail(x, 1)), tail(x,1), col = chart.col, pch = 19, lwd = 5)
  abline(v = as.Date(paste0(seq( year(index(x[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
  misc.NBER.Recessions()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  text(index(tail(x,round(.15*length(x)))[1]), 1.1*tail(x,1), paste("Latest obs.\n", round(tail(x,1),2)), font = 2)
  
  grid()
}




Chart.Four <- function(series1, series2, series3, series4, periods, tweet = FALSE, tweet.text = ""){
  
  x1 = misc.FREDdowload(series = series1)
  x2 = misc.FREDdowload(series = series2)
  x3 = misc.FREDdowload(series = series3)
  x4 = misc.FREDdowload(series = series4)
  
  x1 = x1[year(index(x1))>(year(Sys.Date())-periods)]
  x2 = x2[year(index(x2))>(year(Sys.Date())-periods)]
  x3 = x3[year(index(x3))>(year(Sys.Date())-periods)]
  x4 = x4[year(index(x4))>(year(Sys.Date())-periods)]
  
  par(mfrow=c(2,2))
  
  Chart.Panel(x = x1, series = series1)
  Chart.Panel(x = x2, series = series2)
  Chart.Panel(x = x3, series = series3)
  Chart.Panel(x = x4, series = series4)

  par(mfrow=c(1,1))
}


Chart.InflationOverview <- function(series,tweet = FALSE, tweet.text = "") {
  
  x = data.frame(date = index(get(series[1])),
                 Reduce(function(...) merge(..., all=T), mget(series, ifnotfound = "not found", inherits = TRUE)),
                 year = year(index(get(series[1]))) )
  
  x = aggregate(x, list(x$year), FUN = mean)
  x = x[x$year %in% c("2017", "2018", "2021", "2022"),]
  
  plot.data   =  t( as.matrix(x[3:(ncol(x)-1)]) )
  colnames(plot.data) = x[,"year"]
  plot.ylim   = c(1.15 * min(0, plot.data), 1.15*range(plot.data)[2])
  plot.col    = brewer.pal(nrow(plot.data), "Paired")
  plot.legend = character()
  for (idx in 1 : length(series)) plot.legend = c(plot.legend, gsub("Consumer Price Index for All Urban Consumers: ", "", Data.Description[Data.Description$Mnemonic == series[idx],3]))
  
  bp = barplot(plot.data, beside = T, col = plot.col, main = "Consumer Price Index for All Urban Consumers:\n Average Annual Inflation by Component", ylim = plot.ylim )
  
  limits = par('usr')
  rect(mean(c(bp[nrow(bp),2], bp[1,3])), 0, limits[2], limits[4], lty = 3, lwd = 4)
  
  text(bp, 1.1* plot.data, paste0(round(plot.data,1),"%"), font = 2, cex = .8)
  
  legend("topleft", plot.legend, fill = plot.col, cex = .7, bty = 'n')
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
 
}


misc.GDPExpansionPlot <- function( series ){
  
    x = misc.FREDdowload(series = series)
    
    NBER.Recessions <- as.Date(as.yearqtr(as.Date(as.character(t(nberDates()[,1])), format="%Y%m%d")))
    NBER.Recessions <- NBER.Recessions[year(NBER.Recessions) > 1958]
    
    x <- data.frame( x, StartExpansion = 0)
    
    idx.Expansion                   <- as.Date(rownames(x)) %in% (NBER.Recessions %m-% months(3))
    x$StartExpansion[idx.Expansion] <- 1
    x$StartExpansion                <- cumsum(x$StartExpansion)
    
    if (series == "US.Payroll") col.plot = brewer.pal((max(x$StartExpansion)+1), "Paired") else col.plot = c( brewer.pal((max(x$StartExpansion)+1), "Paired"), "black")
    type.plot    <- c(rep(1,max(x$StartExpansion)),2,2)
    width.plot   <- c(rep(2,max(x$StartExpansion)),5,1)
    legend.plot  <- character()
    
    # Find min and max values for the plot
    plot.ylim.min <- 100
    plot.ylim.max <- 100
    for (idx.plot in 1:max(x$StartExpansion)) {
      tmp.plot <- x[x$StartExpansion == idx.plot,]
      tmp.plot[,1] <- 100 * ( tmp.plot[,1] / tmp.plot[1,1] )
      if (min(tmp.plot[,1], na.rm=TRUE) < plot.ylim.min ) plot.ylim.min <- min(tmp.plot[,1], na.rm=TRUE)
      if (max(tmp.plot[,1], na.rm=TRUE) > plot.ylim.max ) plot.ylim.max <- max(tmp.plot[,1], na.rm=TRUE)
    }
    
    if (series == "US.Payroll") {
      xlim.plot  <- c(1, 120)
      xlim.label = "Months Since Pre-Recession Peak (source: FRED, own calculations)"
    } else { 
      xlim.plot  <- c(1, 40) 
      xlim.label = "Quarters Since Pre-Recession Peak (source: FRED, own calculations)"
      }
    
    ylim.plot    <- c(plot.ylim.min, plot.ylim.max)
    
    tmp.plot <- x[x$StartExpansion == 1,]
    
    legend.plot  <- c(legend.plot, format(as.Date(rownames(tmp.plot)), format = "%B %Y")[2])
    tmp.plot[,1] <- 100 * ( tmp.plot[,1] / as.numeric(tmp.plot[1,1] ))
    
    plot(zoo(tmp.plot[,1]), type="l", lwd=2, col=col.plot[1], lty=type.plot[1], cex.main = .8,
         main=paste0("Comparing Recoveries: \n", Data.Description[Data.Description$Mnemonic == series,3]), 
         ylab="Cumulative Growth (in %) Since Recession", xlab=xlim.label,
         ylim=ylim.plot,
         xlim=xlim.plot,
    )
    
    for (idx.plot in 2:max(x$StartExpansion)) {
      tmp.plot <- x[x$StartExpansion == idx.plot,]
      
      legend.plot <- c(legend.plot, format(as.Date(rownames(tmp.plot)), format = "%B %Y")[2])
      tmp.plot[,1] <- 100 * ( tmp.plot[,1] / as.numeric(tmp.plot[1,1] ))
      lines(zoo(tmp.plot[,1]), lwd=width.plot, col=col.plot[idx.plot], lty=type.plot[idx.plot-3])
    }
    
    
    points( nrow(tmp.plot),      tail(tmp.plot,1)[1], col = "black", pch = 19, lwd = 3)
    if (series == "US.GDP.Real" ) text(   nrow(tmp.plot),  1.2*tail(tmp.plot,1)[1], paste0("After ", nrow(tmp.plot), " Quarters, the Latest Obs. Stands at \n", round(tail(tmp.plot,1)[1],1), "% of the Pre-Recession Peak"), font = 2)
    if (series == "US.Payroll" ) text(   nrow(tmp.plot),  1.2*tail(tmp.plot,1)[1], paste0("After ", nrow(tmp.plot), " Months, the Latest Obs. Stands at \n", round(tail(tmp.plot,1)[1],1), "% of the Pre-Recession Peak"), font = 2)
    
    
    abline(h=100, col="black", lty=3)
    grid()
    
    if (!(series == "US.Payroll")) {
      plot.trend <- 100*(1.03^(1/4))^seq(0, 40)
      lines(plot.trend, lty=3, lwd=3, col=last(col.plot)) 
      legend.plot <- c(legend.plot, "3% Trendline")
    }
    
    legend("topleft", legend.plot, fill=col.plot, cex=0.75, bty = "n")
    legend("bottomright", "Data: St. Louis FRED", cex=.5)
  
}


```

# Backdrop

Every recession is different, but two exceptional ones were the 2008/09 financial crisis and the 2020 COVID pandemic. Both recessions were more severe than any other post-WW II recessions.

How does the post-recession recovery compare to history? To answer this, we index key economic indicators to their pre-recession peak, and trace their evolution through time. 

# Analysis 

Let's look at GDP first, the most comprehensive measure of economi activity. The chart below shows

* the unprecedented depth of the economic contraction, and
* the swift initial rebound, as economic activity restarted,
* the slower expansion as the pandemic continued to linger


```{r GDP Expansion Plot, echo=FALSE}

  misc.GDPExpansionPlot(series = "US.GDP.Real")

```

# Why Does It Matter?

The panels below show the key GDP components. While private consumption is roughly on par with other recoveries, government spending seems a little slower. Overall, given the depths of the recession, the recovery is proceeding remarkably swift so far.



```{r Expansion Details, echo=FALSE}

  par(mfrow=c(2,2))
  misc.GDPExpansionPlot(series = "US.GDP.PCE")
  misc.GDPExpansionPlot(series = "US.GDP.FinalSales")
  misc.GDPExpansionPlot(series = "US.GDP.Investment")
  misc.GDPExpansionPlot(series = "US.GDP.Government")
  par(mfrow=c(1,1))

```

