---
title: "Mortgage Rates"
description: |
  Why we care about mortgage costs
author:
  - name: Philipp Maier
    url: {}
date: 2021-11-12
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=15, fig.height=10) 

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop("Package not found")
  }
}

pkgTest("fredr")
pkgTest("RColorBrewer")
pkgTest("reshape")
pkgTest("quantmod")
pkgTest("zoo")
pkgTest("dplyr")
pkgTest("forecast")
pkgTest("lubridate")
pkgTest("httr")
pkgTest("jsonlite")
pkgTest("rtweet")
pkgTest("httr")
pkgTest("tis")
pkgTest("tm")

## Credentials

FRED_API_KEY <- "4a91875abccae0ee89966c96d8760b60"
fredr_set_key(FRED_API_KEY)


# Data Download

Data.US <- c( "US.Housing.30YMortgageRate"="MORTGAGE30US",
             "US.Housing.15YMortgageRate"="MORTGAGE15US",
             "US.Housing.51ARMMortgageRate"="MORTGAGE5US",
             "US.Housing.30YJumboMortgageRate"="OBMMIJUMBO30YF")


Data.Description <- data.frame( 
  Mnemonic = names(Data.US), 
  Code = Data.US,
  Description = "", 
  Frequeny = "",
  Units = "",
  LastUpdate = ""
  )
  
for ( idx in 1:nrow(Data.Description) ) {
  cat(paste("  - ", Data.Description[idx,1], "\n"))
  x = fredr_series(series_id = Data.Description[idx,2])
  Data.Description[idx, 3 ] = x$title
  Data.Description[idx, 4 ] = x$frequency
  Data.Description[idx, 5] = x$units
  Data.Description[idx, 6] = x$last_updated
}


## Misc. functions

misc.FREDdowload <- function(series) {
  
  if (!(exists(series))) {
    x = fredr( series_id = Data.Description[Data.Description$Mnemonic == series,2] )
    x = zoo(x$value, as.Date(x$date))
    if ((series == "US.H8.FedFundsSold") || (series == "US.H8.TradingAssets")) x = x / 1000
    if ( substr(series, (nchar(series)-5), nchar(series)) == ".mm.yy") x <- 100*log(x / stats::lag(x, -12))
    if ( substr(series, (nchar(series)-5), nchar(series)) == ".qq.yy") x <- 100*log(x / stats::lag(x, -4))
    if ( substr(series, (nchar(series)-2), nchar(series)) == ".qq") x <- 100*log(x / stats::lag(x, -1))
    
  } else assign("x", get(series))
  
  assign(series, x, envir = .GlobalEnv)
  return(x)
}


misc.NBER.Recessions <- function(){
  NBER.Recessions <- as.Date(as.character(t(nberDates())), format="%Y%m%d")
  Limits <- par('usr')
  for (idx in seq(1, length(NBER.Recessions), 2) ) {
    rect(NBER.Recessions[idx], Limits[3], NBER.Recessions[idx+1], Limits[4], col="#0000FF19", lty=0)
  }
}


Chart.Panel <- function(x, series) {
  
  chart.col   = brewer.pal(6, "Paired")[6]
  chart.ylim  = c(0.9*range(x, na.rm=TRUE)[1], 1.1*range(x, na.rm=TRUE)[2])
  chart.title = Data.Description[Data.Description$Mnemonic == series,3]
  
  plot(as.zoo(x), main = chart.title, 
       col = chart.col, 
       lwd = 3, 
       ylim = chart.ylim,
       ylab = Data.Description[grep(series, Data.Description[,1]),5],
       xlab = paste0("Period: ", year(index(x[1])), " - ", year(index(tail(x[1]))), " (shaded areas indicate U.S. recessions)"))
  
  points( index(tail(x, 1)), tail(x,1), col = chart.col, pch = 19, lwd = 5)
  abline(v = as.Date(paste0(seq( year(index(x[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
  misc.NBER.Recessions()
  
  text(index(tail(x,round(.15*length(x)))[1]), 1.1*tail(x,1), paste("Latest obs.\n", round(tail(x,1),2)), font = 2)
  grid()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  
}


Chart.Single <- function(series, periods, tweet = FALSE, tweet.text = ""){
  
  x = misc.FREDdowload(series = series)
  x = x[year(index(x))>(year(Sys.Date())-periods)]
  
  chart.col   = brewer.pal(6, "Paired")[6]
  chart.ylim  = c(0.9*range(x, na.rm=TRUE)[1], 1.1*range(x, na.rm=TRUE)[2])
  chart.title = Data.Description[Data.Description$Mnemonic == series,3]
  if (nchar(chart.title) > 25) chart.title = gsub(":", ":\n", chart.title)
  
  plot(as.zoo(x), main = chart.title, 
       #ylab = "", 
       col = chart.col, 
       lwd = 3, 
       ylim = chart.ylim,
       ylab = "Source: FRED, Federal Reserve Bank of St. Louis",
       xlab = paste0("Period: ", year(index(x[1])), " - ", year(index(tail(x,1))), " (shaded areas indicate U.S. recessions)"))
  points( index(tail(x, 1)), tail(x,1), col = chart.col, pch = 19, lwd = 5)
  abline(v = as.Date(paste0(seq( year(index(x[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
  misc.NBER.Recessions()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  text(index(tail(x,round(.15*length(x)))[1]), 1.1*tail(x,1), paste("Latest obs.\n", round(tail(x,1),2)), font = 2)
  
  grid()
}




Chart.Four <- function(series1, series2, series3, series4, periods, tweet = FALSE, tweet.text = ""){
  
  x1 = misc.FREDdowload(series = series1)
  x2 = misc.FREDdowload(series = series2)
  x3 = misc.FREDdowload(series = series3)
  x4 = misc.FREDdowload(series = series4)
  
  x1 = x1[year(index(x1))>(year(Sys.Date())-periods)]
  x2 = x2[year(index(x2))>(year(Sys.Date())-periods)]
  x3 = x3[year(index(x3))>(year(Sys.Date())-periods)]
  x4 = x4[year(index(x4))>(year(Sys.Date())-periods)]
  
  par(mfrow=c(2,2))
  
  Chart.Panel(x = x1, series = series1)
  Chart.Panel(x = x2, series = series2)
  Chart.Panel(x = x3, series = series3)
  Chart.Panel(x = x4, series = series4)

  par(mfrow=c(1,1))
}

```

# Backdrop

A mortgages is " "a loan used either by purchasers of real property to raise funds to buy real estate, or alternatively by existing property owners to raise funds for any purpose while putting a lien on the property being mortgaged" (Wikipedia). 

About 90% of all homebuyers choose 30-year fixed-rate loans, i.e. a mortgage whereby the interest rate is locked in for 30 years. Other popular choices are 15-year fixed-rate loans, or loans with adjustable rates, whereby rates can change once every 5 years (also known as "5/1 ARM").

# Analysis

The evolution of mortgage rates is worth tracking for several reasons:

* Rising mortgage rates generally increase the cost of purchasing a home, so rising rates tend to slow down the housing market (which represents about 15-18% of US GDP).

* Mortgage rates also offer clues about expectations for future interest rates; all else equal, an increase in the 30-year mortgage rate, relative to the 15-year mortgage rate, may signal that investors tend to expect rates to increase. 

Let's look at some numbers.

```{r analysis}
Chart.Four(series1="US.Housing.30YMortgageRate",
           series2="US.Housing.15YMortgageRate",
             series3="US.Housing.51ARMMortgageRate",
             series4="US.Housing.30YJumboMortgageRate",
             periods = 35,
             tweet = FALSE,
             tweet.text = "")

```

# Why It Matters

Roughly 60% of homes purchased in the US are financed through conventional mortgages. The importance of the mortgage market becomes apparent when realizing that total mortgage debt outstanding in the United States in early 2021 was just under 17 trillion US dollar, or roughly 80% of US GDP (about 21 trillion US dollar). 

Over the coming months, mortgage rates can be expected to rise, as central banks normalize their monetary policy stance. Very rapid rate increases could be disruptive to the housing market, and may hurt affordability.

Updates to these charts will be posted on Twitter as new data becomes available.
