---
title: "COVID19 Update"
description: 
  Why We Care About COVID 19 (a lot!)
author:
  - name: Philipp Maier
date: 2021-11-10
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=15, fig.height=10) 

library(RColorBrewer)
library(COVID19)
library(zoo)
library(lubridate)
library(reshape)

```

# Backdrop

The COVID-19 pandemic has severely disrupted the Global Economy. From lockdowns to bottlenecks in logistics, the pandemic has impacted the economic situation. 

# Analysis

We first look at current infection, i.e. the evolution of daily COVID cases - and, given that this series can be very noise, we also plot the 7-day average. Charts below.

```{r analysis}

covid.countries = c("United States", "China", "Japan", "Germany", "United Kingdom", "India", "France")
covid.cases     = covid19(covid.countries, verbose = FALSE)
plot.col        = brewer.pal(10, "Paired")

par(mfrow = c(2,3))
covid.countries = c("United States", "China", "Japan", "Germany", "United Kingdom", "India")
for (idx in covid.countries) {
  x = covid.cases[covid.cases$administrative_area_level_1==idx, c("date", "confirmed")]
  x = zoo(diff(x$confirmed), as.Date(x$date))
  plot(x, ylab = "Daily Cases (blue) / 7-day average (red)", main = idx, 
       xlab = "", xaxt = "n", col = plot.col[1], lwd = 3, cex.main = 2.2, cex.axis = 1.4, cex.lab = 1.3)
  lines(rollmean(x, k=7), col = plot.col[6], lwd = 3)
  axis(1, floor_date(index(x), 'month'), as.yearmon(floor_date(index(x), 'month')), cex.axis = 1)
#  legend("topleft", "Data: https://covid19datahub.io", cex=1.2, bty = "n")
  grid()
}
par(mfrow = c(1,1))

```


# Why It Matters

One notable impact is the decline in the labor participation rate, which could potentially trigger a slowdown in productivity and drive up inflation. This underlines the importance to monitor COVID-19 trends across countries.

What's the best way to avoid getting infected? High vaccination rates, as vaccines provide the best long-term protection. We track vaccination rates in the following chart (note that China is not included, as the vaccination data in the data set does not look credible yet).


```{r whyitmatters}

covid.countries = c("United States", "Japan", "Germany", "United Kingdom", "India", "France")

x = covid.cases[, c("date", "administrative_area_level_1", "people_vaccinated", "population")]
x$vaccinated_ratio = x$people_vaccinated/x$population*100

y = melt(x, id=c("date", "administrative_area_level_1") )
x = cast(y[y$variable=="vaccinated_ratio",], date ~ administrative_area_level_1 + variable)
x = zoo(x, as.Date(x$date))
colnames(x) = gsub( "_vaccinated_ratio","", colnames(x))
x= x[index(x) > as.Date("2020-12-01"),covid.countries]

plot(x[,1], type = "n", ylab = "% of population", main = "Vaccinated as % of population", ylim = c(0,100),
     xlab = "", 
     xaxt = "n",
     xlim = c(head(index(x),1), tail(index(x),1))  )
axis(1, seq( head(index(x),1), tail(index(x),1) , "months" ),
     as.character( as.yearmon( seq( head(index(x),1), tail(index(x),1), "months" ))), cex.axis = .8)

for (idx in 1:length(covid.countries)) {
  lines( x[,idx], col = plot.col[idx], lwd = 3)
  points(index(tail(na.omit(x[,idx]),1)), tail(na.omit(x[,idx]),1), pch = 19, lwd = 5, col = plot.col[idx])
}

legend("topleft", colnames(x), fill = plot.col, bty = "n")
#legend("bottomright", "Data: https://covid19datahub.io", bty = "n")
grid()


```

Updates to these charts will be posted on Twitter as new data becomes available.


## A note on data

* All data from the COVID-19 Data Hub (https://covid19datahub.io). 

* Data is accessed via the terrific COVID19 R-package, about which you can find more information in Guidotti, E., Ardia, D., (2020), "COVID-19 Data Hub", Journal of Open Source Software 5(51):2376, https://doi.org/10.21105/joss.02376

 
