---
title: "Shifting Consumption Patterns"
description: 
  With Travel Restricted, Consumers Continue Their Serious Retail Therapy. What Does It Mean for the Economy?
author:
  - name: Philipp Maier
    url: https://twitter.com/EconomicsShiny
date: 2021-11-07
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

knitr::opts_chunk$set(fig.width=15, fig.height=10) 

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop("Package not found")
  }
}

pkgTest("fredr")
pkgTest("RColorBrewer")
pkgTest("reshape")
pkgTest("quantmod")
pkgTest("zoo")
pkgTest("dplyr")
pkgTest("forecast")
pkgTest("lubridate")
pkgTest("httr")
pkgTest("jsonlite")
pkgTest("rtweet")
pkgTest("httr")
pkgTest("tis")
pkgTest("tm")

## Credentials

FRED_API_KEY <- "4a91875abccae0ee89966c96d8760b60"
fredr_set_key(FRED_API_KEY)



# Data Download

Data.US <- c("US.Activity.RetailSales" = "RSAFS",
             "US.Activity.RetailSalesExAuto" = "RSFSXMV",
             "US.Activity.RetailSalesMotorvehicles" = "RSMVPD",
             "US.Activity.RetailSalesElectronics" = "RSEAS",
             "US.Activity.RetailSalesFoodDBeverageRetail" = "RSDBS",
             "US.Activity.RetailSalesHealth" = "RSHPCS",
             "US.Activity.RetailSalesGasoline" = "RSGASS",
             "US.Activity.RetailSalesGeneralMerchandise" = "RSGMS",
             "US.Activity.RetailSalesMisc" = "RSMSR",
             "US.Activity.RetailSalesNonStore" = "RSNSR",
             "US.Activity.RetailSalesSportsGoods" = "RSSGHBMS",
             "US.Activity.RetailSalesFurniture" = "RSFHFS",
             "US.Activity.RetailSalesBuildingMaterials" = "RSBMGESD",
             "US.Activity.RetailSalesClothing" = "RSCCAS",
             "US.Activity.RetailSalesFoodDrinking" = "RSFSDP"
)


Data.Description <- data.frame( 
  Mnemonic = names(Data.US), 
  Code = Data.US,
  Description = "", 
  Frequeny = "",
  Units = "",
  LastUpdate = ""
  )
  
for ( idx in 1:nrow(Data.Description) ) {
  cat(paste("  - ", Data.Description[idx,1], "\n"))
  x = fredr_series(series_id = Data.Description[idx,2])
  Data.Description[idx, 3 ] = x$title
  Data.Description[idx, 4 ] = x$frequency
  Data.Description[idx, 5] = x$units
  Data.Description[idx, 6] = x$last_updated
}


## Misc. functions

misc.FREDdowload <- function(series) {
  
  if (!(exists(series))) {
    x = fredr( series_id = Data.Description[Data.Description$Mnemonic == series,2] )
    x = zoo(x$value, as.Date(x$date))
    if ((series == "US.H8.FedFundsSold") || (series == "US.H8.TradingAssets")) x = x / 1000
    if ( substr(series, (nchar(series)-5), nchar(series)) == ".mm.yy") x <- 100*log(x / stats::lag(x, -12))
    if ( substr(series, (nchar(series)-5), nchar(series)) == ".qq.yy") x <- 100*log(x / stats::lag(x, -4))
    if ( substr(series, (nchar(series)-2), nchar(series)) == ".qq") x <- 100*log(x / stats::lag(x, -1))
    
  } else {
    
    assign("x", get(series))
  }
  
  assign(series, x, envir = .GlobalEnv)
  return(x)
}


misc.NBER.Recessions <- function(){
  NBER.Recessions <- as.Date(as.character(t(nberDates())), format="%Y%m%d")
  Limits <- par('usr')
  for (idx in seq(1, length(NBER.Recessions), 2) ) {
    rect(NBER.Recessions[idx], Limits[3], NBER.Recessions[idx+1], Limits[4], col="#0000FF19", lty=0)
  }
}



Chart.Panel <- function(x, series) {
  
  chart.col   = brewer.pal(6, "Paired")[6]
  chart.ylim  = c(0.9*range(x, na.rm=TRUE)[1], 1.1*range(x, na.rm=TRUE)[2])
  chart.title = Data.Description[Data.Description$Mnemonic == series,3]
  
  plot(as.zoo(x), main = chart.title, 
       col = chart.col, 
       lwd = 3, 
       ylim = chart.ylim,
       ylab = Data.Description[grep(series, Data.Description[,1]),5],
       xlab = paste0("Period: ", year(index(x[1])), " - ", year(index(tail(x[1]))), " (shaded areas indicate U.S. recessions)"))
  
  points( index(tail(x, 1)), tail(x,1), col = chart.col, pch = 19, lwd = 5)
  abline(v = as.Date(paste0(seq( year(index(x[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
  misc.NBER.Recessions()
  
  text(index(tail(x,round(.15*length(x)))[1]), 1.1*tail(x,1), paste("Latest obs.\n", round(tail(x,1),2)), font = 2)
  grid()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  
}


Chart.Single <- function(series, periods, tweet = FALSE, tweet.text = ""){
  
  x = misc.FREDdowload(series = series)
  x = x[year(index(x))>(year(Sys.Date())-periods)]
  
  chart.col   = brewer.pal(6, "Paired")[6]
  chart.ylim  = c(0.9*range(x, na.rm=TRUE)[1], 1.1*range(x, na.rm=TRUE)[2])
  chart.title = Data.Description[Data.Description$Mnemonic == series,3]
  if (nchar(chart.title) > 25) chart.title = gsub(":", ":\n", chart.title)
  
  plot(as.zoo(x), main = chart.title, 
       #ylab = "", 
       col = chart.col, 
       lwd = 3, 
       ylim = chart.ylim,
       ylab = "Source: FRED, Federal Reserve Bank of St. Louis",
       xlab = paste0("Period: ", year(index(x[1])), " - ", year(index(tail(x,1))), " (shaded areas indicate U.S. recessions)"))
  points( index(tail(x, 1)), tail(x,1), col = chart.col, pch = 19, lwd = 5)
  abline(v = as.Date(paste0(seq( year(index(x[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
  misc.NBER.Recessions()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  text(index(tail(x,round(.15*length(x)))[1]), 1.1*tail(x,1), paste("Latest obs.\n", round(tail(x,1),2)), font = 2)
  
  grid()
}


Chart.Duo <- function(series1, series2, periods, tweet = FALSE, tweet.text = ""){
  
    x1 = misc.FREDdowload(series = series1)
    x2 = misc.FREDdowload(series = series2)
    
    x1 = x1[year(index(x1))>(year(Sys.Date())-periods)]
    x2 = x2[year(index(x2))>(year(Sys.Date())-periods)]
 
    par(mfrow=c(1,2))
    
    chart.col   = brewer.pal(6, "Paired")[6]
    chart.ylim  = c(0.9*range(x1, na.rm=TRUE)[1], 1.1*range(x1, na.rm=TRUE)[2])
    chart.title = Data.Description[Data.Description$Mnemonic == series1,3]
    
    plot(as.zoo(x1), main = chart.title, 
         col = chart.col, 
         lwd = 3, 
         ylim = chart.ylim,
         #ylab = "Source: FRED, Federal Reserve Bank of St. Louis",
         ylab = ifelse ( substr(series1, (nchar(series1)-5), nchar(series1)) == ".qq.yy", "Y/Y growth (in %)", Data.Description[grep(series1, Data.Description[,1]),5]),
         xlab = paste0("Period: ", year(index(x1[1])), " - ", year(index(tail(x1,1))), " (shaded areas indicate U.S. recessions)"))
    points( index(tail(x1, 1)), tail(x1,1), col = chart.col, pch = 19, lwd = 5)
    abline(v = as.Date(paste0(seq( year(index(x1[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
    misc.NBER.Recessions()
    
    text(index(tail(x1,round(.15*length(x1)))[1]), 1.1*tail(x1,1), 
         ifelse ( substr(series1, (nchar(series1)-5), nchar(series1)) == ".qq.yy", 
                  paste("Latest obs.\n", round(tail(x1,1),2),"%"),
                  paste("Latest obs.\n", round(tail(x1,1),2))), font = 2)
    grid()
    legend("bottomleft", "Data: St. Louis FRED", cex=.6)
    
    chart.col   = brewer.pal(6, "Paired")[6]
    chart.ylim  = c(0.9*range(x2, na.rm=TRUE)[1], 1.1*range(x2, na.rm=TRUE)[2])
    chart.title = Data.Description[Data.Description$Mnemonic == series2, 3]
    
    plot(as.zoo(x2), main = chart.title, 
         col = chart.col, 
         lwd = 3, 
         ylim = chart.ylim,
         ylab = ifelse ( substr(series2, (nchar(series2)-5), nchar(series2)) == ".qq.yy", "Y/Y growth (in %)", Data.Description[grep(series2, Data.Description[,1]),5]),
         xlab = paste0("Period: ", year(index(x2[1])), " - ", year(index(tail(x2,1))), " (shaded areas indicate U.S. recessions)"))
    points( index(tail(x2, 1)), tail(x2,1), col = chart.col, pch = 19, lwd = 5)
    abline(v = as.Date(paste0(seq( year(index(x2[1])), year(Sys.Date()), 1), "-01-01")), lty = 3, lwd = 0.5)
    misc.NBER.Recessions()
    
    text(index(tail(x2,round(.15*length(x2)))[1]), 1.1*tail(x2,1), 
         ifelse ( substr(series2, (nchar(series2)-5), nchar(series2)) == ".qq.yy",
                  paste("Latest obs.\n", round(tail(x2,1),2),"%"),
                  paste("Latest obs.\n", round(tail(x2,1),2))), font = 2)
    grid()
    legend("bottomleft", "Data: St. Louis FRED", cex=.6)
    
    par(mfrow=c(1,1))
   
}



Chart.Four <- function(series1, series2, series3, series4, periods, tweet = FALSE, tweet.text = ""){
  
  x1 = misc.FREDdowload(series = series1)
  x2 = misc.FREDdowload(series = series2)
  x3 = misc.FREDdowload(series = series3)
  x4 = misc.FREDdowload(series = series4)
  
  x1 = x1[year(index(x1))>(year(Sys.Date())-periods)]
  x2 = x2[year(index(x2))>(year(Sys.Date())-periods)]
  x3 = x3[year(index(x3))>(year(Sys.Date())-periods)]
  x4 = x4[year(index(x4))>(year(Sys.Date())-periods)]
  
  par(mfrow=c(2,2))
  
  Chart.Panel(x = x1, series = series1)
  Chart.Panel(x = x2, series = series2)
  Chart.Panel(x = x3, series = series3)
  Chart.Panel(x = x4, series = series4)

  par(mfrow=c(1,1))
}


Chart.DoublePanel <- function(series1, series2, periods) {
  
  x1 = misc.FREDdowload(series1) 
  x2 = misc.FREDdowload(series2)
  
  plot.data = merge(x1, x2)
  plot.data = plot.data[year(index(plot.data))>(year(Sys.Date())-periods)]
  names(plot.data) <- gsub("Motor Vehicle Retail Sales: ", "", c(Data.Description[Data.Description$Mnemonic == series1,3], Data.Description[Data.Description$Mnemonic == series2,3]) )
  
  chart.col   = brewer.pal(6, "Paired")[1:2]
  chart.ylim  = c(0.9*range(plot.data, na.rm=TRUE)[1], 1.1*range(plot.data, na.rm=TRUE)[2])
  chart.title = "Motor Vehicle Retail Sales" #Data.Description[Data.Description$Mnemonic == series1,3]
  
  plot(plot.data[,1], col = chart.col, main = chart.title, type = "n", 
       ylim = chart.ylim, 
       ylab = Data.Description[Data.Description$Mnemonic == series1,5], 
       xlab = "")
  legend("topleft", names(plot.data), fill = chart.col, bty = "n")
  
  for (idx in 1:ncol(plot.data)) {
    lines(plot.data[,idx], col = chart.col[idx])
    points( index(tail(plot.data[,idx], 1)), tail(plot.data[,idx],1), col = chart.col[idx], pch = 19, lwd = 5)
    text(index(tail(plot.data[,idx],round(.15*length(x2)))[1]), 1.1*tail(plot.data[,idx],1), 
         paste("Latest obs.\n", round(tail(plot.data[,idx],1),2)), font = 2)
  }
  misc.NBER.Recessions()
  
  grid()
  legend("bottomleft", "Data: St. Louis FRED", cex=.5)
  
}



```


# Backdrop

The pandemic, and the closure of cross-border travel, resulted in substantial disruptions for the hospitality and travel industry. Partially as a result of government support, consumption remained high, and consumers spent money on their homes - i.e. on goods like furniture, sporting goods, or building materials.

This causes massive surges in demand for these goods, and retailers had difficulties meeting unexpected demands. How long before consumption spending returns to more normal levels?

The chart below illustrates just how disruptive the pandemic was.


```{r Backdrop }

Chart.Duo(series1="US.Activity.RetailSales", series2="US.Activity.RetailSalesExAuto",
          periods = 25,
          tweet = ( as.Date( as.character(tail( fredr_release_dates(release_id = 9L) ,1)[,2]) ) == Sys.Date() ),  
          tweet.text = "#AdvanceRetailSales: With Travel Restricted, Consumers Continue Their Serious #RetailTherapy #rstats")

```

# Analysis 

Let's dig below the headline number. The chart below shows key components from the monthly Advance Retail Sales Report, released monthly by the U.S. Census Bureau (https://fred.stlouisfed.org/release/tables?rid=9&eid=201241#snid=201244). 

The second chart shows the impact of the pandemic even more clearly. Bought some exercise equipment lately? You're not alone.




```{r Analysis }


BankData.Details  <- merge(US.Activity.RetailSalesBuildingMaterials=misc.FREDdowload("US.Activity.RetailSalesBuildingMaterials"),
                           #US.Activity.RetailSalesMisc=misc.FREDdowload("US.Activity.RetailSalesMisc"),
                           US.Activity.RetailSalesGeneralMerchandise=misc.FREDdowload("US.Activity.RetailSalesGeneralMerchandise"),
                           US.Activity.RetailSalesFoodDBeverageRetail=misc.FREDdowload("US.Activity.RetailSalesFoodDBeverageRetail"),
                           US.Activity.RetailSalesGasoline=misc.FREDdowload("US.Activity.RetailSalesGasoline"),
                           US.Activity.RetailSalesClothing = misc.FREDdowload("US.Activity.RetailSalesClothing"),
                           US.Activity.RetailSalesMotorvehicles=misc.FREDdowload("US.Activity.RetailSalesMotorvehicles"),
                           US.Activity.RetailSalesNonStore=misc.FREDdowload("US.Activity.RetailSalesNonStore"),
                           US.Activity.RetailSalesFoodDrinking = misc.FREDdowload("US.Activity.RetailSalesFoodDrinking")
)

BankData.Details  <- BankData.Details[index(BankData.Details) >= as.Date("2019-01-01")] / 1000

plot.col    <- brewer.pal(8, "Paired")
plot.legend = character()
for (idx in 1:ncol(BankData.Details)) plot.legend <- c(plot.legend, 
                                                       gsub("Advance Retail Sales: ", "", 
                                                            Data.Description[Data.Description$Mnemonic == colnames(BankData.Details)[idx],3]))
plot.ylim = c(0, 1.25*max(rowSums(BankData.Details)))
BankDataChart=barplot(BankData.Details, col = plot.col, 
                      border="NA", 
                      main = "Advance Retail Sales: Key Components", 
                      ylab = "Billions of Dollars",
                      xaxt = "n",
                      cex.axis = .8, cex.names = .8,
                      ylim = plot.ylim)
axis(1, at=BankDataChart,
     as.yearmon( index(BankData.Details)) , cex.axis = .7)
legend("topleft", plot.legend, fill = plot.col, cex=0.75, bty = "n")
legend("bottomleft", "Data: St. Louis FRED", cex=.5)



Chart.Duo(series1="US.Activity.RetailSalesSportsGoods", series2="US.Activity.RetailSalesFurniture",
          periods = 25,
          tweet = ( as.Date( as.character(tail( fredr_release_dates(release_id = 9L) ,1)[,2]) ) == Sys.Date() ),  
          tweet.text = "Bought a #Peleton During the Pandemic, too? Sales of #SportingGoods Skyrocketed in 2021.
           (source: FRED) #AdvanceRetailSales #rstats")




```


Updates to these charts will be posted on Twitter as new data becomes available.
